# (PART\*) Functional Proteomics 

```{r include=FALSE}
library(pOmics3)
```

# Introduction




# TPP




# SPP 




# PELSA


## The HT.PELSA package



## QC for Dose-Response Results


Let's load the libraries first. 

```{r eval=FALSE}
library(tidyverse)
library(pOmics3)
library(UpSetR)
library(ComplexHeatmap)
```


The PELSA output from the `HT.PELSA` package does not immediately show in which replicates a peptide was identified as a hit. This information can be retrieved from the `data_EC50$data_replicates` table. 

```{r eval=FALSE}
# Extract replicate hits per peptide 
data_hit_replicates <- data_EC50$data_replicates %>% 
  filter(regulation %in% c("stabilized", "destabilized") & 
           regulation_replicate %in% c("stabilized", "destabilized")) %>% 
  arrange(Peptides, Replicate) %>% 
  summarise(Replicate.Hits = 
              paste(Replicate[regulation %in% c("stabilized", "destabilized")], 
                                   collapse = "/"), 
            .by = "Peptides")
```



Next, we can look at the distribution of peptide hits in across the different replicates.

```{r eval=FALSE}
# Peptide hit distribution over replicates 
data_EC50$data_replicates %>% 
  # Identify hits 
  filter(regulation %in% c("stabilized", "destabilized")) %>% 
  # Remove unchanged peptide replicates 
  filter(regulation_replicate %in% c("stabilized", "destabilized")) %>% 
  arrange(Peptides, Replicate) %>% 
  pivot_wider(id_cols = "Peptides", 
              names_from = "Replicate", 
              values_from = "regulation") %>% 
  mutate(across(2:5, \(x) as.numeric(!is.na(x)))) %>% 
  pOmics3::t2df() %>% 
  UpSetR::upset(order.by = "freq", 
                mainbar.y.label = "# peptide hits", 
                sets.x.label = "# peptide hits")
```

To investigate, whether the the identification biased the replicates towards different numbers of hits, we can check in how many replicates each hit was quantified and a curve fitted.

```{r eval=FALSE}
# Distribution of fitted curves for hits 
data_EC50$data_replicates %>% 
  # Identify hits 
  dplyr::filter(regulation %in% c("stabilized", "destabilized")) %>% 
  dplyr::arrange(Peptides, Replicate) %>% 
  tidyr::pivot_wider(id_cols = "Peptides", 
              names_from = "Replicate", 
              values_from = "do.fit") %>% 
  dplyr::mutate(across(2:5, \(x) as.numeric(tidyr::replace_na(x, F)))) %>% 
  pOmics3::t2df() %>% 
  UpSetR::upset(order.by = "freq", 
                mainbar.y.label = "# fitted peptides curves (among hits)", 
                sets.x.label = "(same as y-axis)")
```

Looking good! 

Considering now all peptides for which curves were fitted, we can visualize the first upset plot but this with all fitted peptide curves, independent of them being identified as a hit in the end. 

```{r eval=FALSE}
# Hit distribution of all fitted curves 
data_EC50$data_replicates %>% 
  #filter(regulation %in% c("stabilized", "destabilized")) %>% 
  dplyr::filter(regulation_replicate %in% c("stabilized", "destabilized")) %>% 
  dplyr::arrange(Peptides, Replicate) %>% 
  tidyr::pivot_wider(id_cols = "Peptides", 
              names_from = "Replicate", 
              values_from = "regulation") %>% 
  mutate(across(2:5, \(x) as.numeric(!is.na(x)))) %>% 
  pOmics3::t2df() %>% 
  UpSetR::upset(order.by = "freq", 
                mainbar.y.label = "# peptide replicate hits", 
                sets.x.label = "(same as y-axis)")
```


Finally, we can see if the number of fitted peptide curves was equal among all replicates for all peptides.

```{r eval=FALSE}
# Distribution of fitted curves for hits 
data_EC50$data_replicates %>% 
  #filter(regulation %in% c("stabilized", "destabilized")) %>% 
  dplyr::arrange(Peptides, Replicate) %>% 
  tidyr::pivot_wider(id_cols = "Peptides", 
              names_from = "Replicate", 
              values_from = "do.fit") %>% 
  dplyr::mutate(across(2:5, \(x) as.numeric(tidyr::replace_na(x, F)))) %>% 
  pOmics3::t2df() %>% 
  UpSetR::upset(order.by = "freq", 
                mainbar.y.label = "# fitted peptides curves (all peptides)", 
                sets.x.label = "(same as y-axis)")
```


Further, we can count the number of peptides quantified per replicate and concentration, and represent it in a heatmap. 

```{r eval=FALSE}
# Heatmap of number of quantified peptides 
data_EC50$data_replicates %>% 
  dplyr::select(Peptides, Replicate, found_in) %>% 
  tidyr::separate_longer_delim(found_in, "/") %>% 
  dplyr::count(Replicate, found_in) %>% 
  dplyr::filter(found_in!= "") %>% 
  tidyr::pivot_wider(names_from = "found_in", values_from = n) %>% 
  t2m() %>% 
  #t() %>% 
  ComplexHeatmap::Heatmap(cluster_columns = F, name = "# quantified peptides")
```

Don't get fooled by the automatic color scale!

Sometimes, there may be a bias within the EC50 values. We can also visualize this.

```{r eval=FALSE}
# Heatmap of pEC50 values for hits 
data_EC50$data_replicates %>% 
  dplyr::filter(!is.na(pEC50_mean)) %>% 
  tidyr::pivot_wider(id_cols = "Peptides", 
                     names_from = "Replicate", 
                     values_from = "pEC50") %>% 
  t2m() %>% 
  ComplexHeatmap::Heatmap(name = "pEC50", 
                          show_row_names = F)
```

This can be separated into stabilized and destabilized peptides as well.

`r button_begin("stabilized")`
\
```{r eval=FALSE}
# for stabilized hits
data_EC50$data_replicates %>% 
  dplyr::filter(!is.na(pEC50_mean) & regulation == "stabilized") %>% 
  tidyr::pivot_wider(id_cols = "Peptides", 
                     names_from = "Replicate", 
                     values_from = "pEC50") %>% 
  t2m() %>% 
  ComplexHeatmap::Heatmap(name = "pEC50", 
                          show_row_names = F)
```
`r button_end()`

`r button_begin("destabilized")`
\
```{r eval=FALSE}
# for destabilized hits
data_EC50$data_replicates %>% 
  dplyr::filter(!is.na(pEC50_mean) & regulation == "destabilized") %>% 
  tidyr::pivot_wider(id_cols = "Peptides", 
                     names_from = "Replicate", 
                     values_from = "pEC50") %>% 
  t2m() %>% 
  ComplexHeatmap::Heatmap(name = "pEC50", 
                          show_row_names = F)
```
`r button_end()`

\
Maybe it helps to plot the difference between the replicate and the mean pEC50 values. This can be done like this.

```{r eval=FALSE}
# pEC50 - pEC50_mean values
data_EC50$data_replicates %>% 
  dplyr::filter(!is.na(pEC50_mean)) %>% 
  dplyr::mutate(pEC50 = pEC50 - pEC50_mean) %>% 
  tidyr::pivot_wider(id_cols = "Peptides", 
                     names_from = "Replicate", 
                     values_from = "pEC50") %>% 
  t2m() %>% 
  ComplexHeatmap::Heatmap(name = "pEC50 - pEC50_mean", 
                          show_row_names = F)
```

This can be further scaled by dividing the difference by the mean value again. 

`r button_begin("scaled")`
\
```{r eval=FALSE}
# Scaled pEC50 - pEC50_mean values
data_EC50$data_replicates %>% 
  dplyr::filter(!is.na(pEC50_mean)) %>% 
  dplyr::mutate(pEC50 = (pEC50 - pEC50_mean) / pEC50_mean) %>% 
  tidyr::pivot_wider(id_cols = "Peptides", 
                     names_from = "Replicate", 
                     values_from = "pEC50") %>% 
  t2m() %>% 
  ComplexHeatmap::Heatmap(name = "(pEC50 - pEC50_mean) / pEC50_mean", 
                          show_row_names = F)
```
`r button_end()`

\
Finally, the pEC50 values can be compared among replicates via a correlation matrix of pEC50 values. 


```{r eval=FALSE}
# Correlation matrix of pEC50 values 
data_EC50$data_replicates %>% 
  dplyr::filter(!is.na(pEC50_mean)) %>% 
  tidyr::pivot_wider(id_cols = "Peptides", 
                     names_from = "Replicate", 
                     values_from = "pEC50") %>% 
  t2m() %>% 
  cor(use = "pairwise.complete.obs") %>% 
  ComplexHeatmap::Heatmap(name = "pEC50")
```

